---
title: "week6notes"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(here)
```

# Monday

## Vectors and Functions

we often want to ‘run’ our function over multiple inputs

- many citizen scientists contributing to species data collection in different locations
- Generally vector arithmetic doesn’t require a for loop (in most languages)
- R has many ways to loop over data

we will cover several
simplest vector arithmetic (built in)

```{r}
# example
a <- c(1, 2, 3)
b <- c(4, 5, 6)
a + b
# sum is done for each element of the vector
```
** Implicit Iteration** = If calculations in the function can be done as vectors - this is automatic (*e.g. parallel rows for each variable*)


Example
```{r}
source(here("week6/power_gen_orig.R"))

power_gen_orig(flow=2,height=5)
```
```{r}
# Vectors of flow and height
flow = runif(min=1, max=10, n=20)
height = runif(min=5, max =30, n=20)

power = power_gen_orig(flow, height)
power

```
```{r}
length(power)
length(flow)
```
```{r}
power_gen_orig(flow=flow, height = height)
```

What if lengths are unequal?? unclear results

```{r}
height = c(0,1)
power = power_gen_orig(flow, height)
# it works but what is it doing?
 power
```
```{r}
# unclear behaviors - best to not do it
# helps to define a data frame
powerdf = data.frame(
flow = runif(min=1, max=10, n=20),
height = runif(min=5, max =30, n=20)
)

# this works
 powerdf$highK = power_gen_orig(flow, height, K=0.4)
 powerdf$lowK = power_gen_orig(flow, height, K=0.9)
 mean(powerdf$lowK)
```
```{r}
power1 = power_gen_orig(flow, height, Keff = 0.1)
power2 = power_gen_orig(flow, height, Keff = 0.8)
```

```{r}
powerdf = data.frame(
  flow = runif(min=1, max = 10, n=20),
  height = runif(min=5, max = 30, n = 20)
)

powerdf$highK = power_gen_orig(flow, height, K=0.4)
```

## Loops and Functions

**Another Example**

Calculate NPV for a range of diff interest rates, a range of damages, that may be incurred 10 years into the future

Steps
- define inputs (interest rates, damages)
- define output (NPV)
- write the function
- create a data structure to store results where we vary both interest rates and damages
- use nested for loops to fill in the data structure

#### NPV: Nested For Loop Example
```{r}
# write a function to compute npv

compute_NPV = function(value, time, discount) {
  #use value to compute NPV
  result = value / (1+discount)**time
  return(result)
}

compute_NPV
```
```{r}
#test the function
compute_NPV(20, discount=0.01, time=20)
```
```{r}
#what we wanted to be able to do, is run a bunch of possible inputs and rates to see what could happen
# generate some input
damages = c(25,33,91,24)
damages

# sensitivity to discount rate
discount_rates = seq(from=0.01, to=0.04, by=0.005)
discount_rates

#let's say we always care about the money in 10 years
yr = 10
```
```{r}
compute_NPV(value = damages, discount = discount_rates, time=yr)

# R doesn't like this bc there are diff number of rates, damages, etc

# we need a for loop!
```
```{r}
npvs = data.frame(
  matrix(nrow=length(damages), ncol=length(discount_rates))
)
# make an empty data frame to initialize for our for loop
```

```{r}
for (i in 1:length(damages)) {
  for (j in 1:length(discount_rates)) {
    npvs[i,j]= compute_NPV(value=damages[i], discount = discount_rates[j], time=yr)
  }
}
npvs

colnames(npvs) = discount_rates
rownames(npvs) = damages
npvs$damages = damages
```
For plotting, pivot_longer is helpful to reorganize
```{r}
npv_plotdata = npvs |> pivot_longer(!damages, names_to = "discount", values_to = "npv" )

# now it's better to plot
```

```{r}
npv_plot <- ggplot(npv_plotdata, aes(x = damages, y = npv, color = discount)) +
  geom_point() +
  theme_minimal()

npv_plot
```

## Thinking Big: how to create modules as parts of a big program

### A Simple Function Example - Just One Module
Let’s say we want to develop a program that computes power generated from a hydro-reservoir

We know that power generation depends on water release rates and the potential energy (from the reservoir height)



### Let’s Think About a More Complex Program…

Goal = Calculate profit from a reservoir fed by a river, if a minimum reservoir volume must be maintained and otherwise the reservoir is operated at a flow rate set by the user

Input =  Stream gage data (m^^3^^/sec)

Parameters =
Efficiency,
ρ (density of water),
g (acceleration due to gravity), 
Maximum Reservoir Capacity (m3), 
Minimum Reservoir Volume (m3), 
Set flow rate (m3/sec), 
Price of power (/W)$, 

Output = Profit ($)

Modules

- Compute Power from a given reservoir height and flow rate (compute_power)
- Keep track of reservoir volume given inflow and output (reservoir_state)
- Determine flow rate, given operator desired flow rate but need for minimum capacity (determine_flow_rate)
- Compute profit from power (compute_profit)
Each module would have its own inputs, parameters and output

A Flow control diagram puts the modules together

### Assignment 5

With a partner, think of an example complex data science task/program that you might want to design with:

- at least 4 modules
- 1 module must be used at least twice
- at least one for (or while) loop

Planning - Step 1

- sketch a diagram that shows how these modules would be put together to go from inputs to output for the overall program (30pts)
- write a contract for the overall program (input, output, parameters) (10pts)
- break the program into modules (naming them appropriately), write a contract for each one (20pts)

*Important - Check in with myself or Ojas to get the OK on your plan BEFORE you go to Step 2*

Implementation - Step 2

- write functions (stored in separate documents) for each module (30pts)
- write an overall program that manages calls to each function (10pts)
- create some data and use it to test the program (10pts)


Ideas

based on amount of precipitation, calculate the height of the water (goes into both options) 

costs for no action (old bridge)
- accidents/fatalities
- infrastructure damages as a function of precipitation
- increased travel times due to rerouting when bridge closed

costs for building new bridge
- construction materials
- construction labor
- if there's too much rain, have to stop construction and delays cost X amount (for loop)

final module 
- add modules of each option and print results


# Tuesday 

## Control Structures

### If Else in R
Two ways in R

**ifelse function**
- ifelse(condition, TRUE, FALSE)
- best used for simple application

**if..else structure**
more formal way of deciding on two paths

```{r}
a = 20
classr  = ifelse(a > 90, "high","low")
classr
```
```{r}
# also works with multiple values
a = runif(min=1, max=100, n=5)
a
```
```{r}
# ifelse can work with vectors
classr  = ifelse((a > 90) | !(a<10), "high","low")
classr
```
:::{.callout-note}
& <-- and

| <-- or

! <-- except
:::

#### Flow control in functions

best to use `if {} else {}`

- see example in `summarize_diversity2.R`

```{r}
source(here("week6/summarize_diversity2.R"))

flowers = sample(c("rose","peony","daisy"), replace=T, prob = c(0.1, 0.8, 0.1),size=30)

summarize_diversity(flowers, index="B")$simpson
```
```{r}
summarize_diversity(flowers, index="A")$simpson
# can specify index A or B so that the function knows which formula to use to calculate biodiversity
```
#### Embedding ifelse statements

Gives us more than two conditions
```{r}
classr= ifelse(a>10, ifelse(a>50, "high", "med"), "low")
```

### Multiple Conditions

If you have multiple conditions we can use `case_when` (requires tidyverse library)

or `switch` is another option

#### `case_when` Fishing Example

Here’s a simple example of how it works

Imagine you are recording fish catches at different locations (we are going to simulate this by sampling )

Fish have grades… A, B, C
- steelhead are A,
- tuna are B
- everything else is C

:::{.callout-note}
case_when(

condition ~ response,

condition ~ response … )
:::

```{r}
library(tidyverse)
fish = "steelhead"
case_when(
  (fish =="steelhead")  ~ "A", 
  (fish =="tuna") ~ "B", 
  !(fish == "tuna") & !(fish=="steelhead") ~ "C"
  )
```
Let's build a function to make this cleaner
```{r}
# apply to all of our ocean
# start by making a function - this will be cleaner!
classify_fish = function(fish) {
 class =  case_when(
   (fish =="steelhead")  ~ "A", 
   (fish =="tuna") ~ "B", 
   !(fish == "tuna") & !(fish=="steelhead") ~ "C")
 return(class) }

classify_fish("steelhead")
classify_fish("cod")
```
```{r}
# apply to our ocean
possible.fish = c("salmon","steelhead","shark","tuna","cod")
catch2 = base::sample(possible.fish, size=20, replace=T)
catch2

# now classify all samples
classify_fish(catch2)
```

#### A more interesting `case_when` example

Lets imagine that we are monitoring pollution in a lake, and we want to write a function that will let us know (flag) if risk associated with nutrient pollution are high, medium or low

From ecological studies, we know that:

*High Risk Conditions*
- risk is high if water temperature is greater than a threshold for more than 5 days, and
- mean nutrient concentration is greater than a high threshold

*Medium Risk Conditions*
- Risk is medium if water temperature is greater than a threshold for more than 5 days and
- mean nutrient concentration is greater than a medium threshold

*Everything else is low risk*

**Inputs**:
- nutrient concentration for at least 5 days
- temperature for at least 5 days
- thresholds for temperature and nutrient (with default values)

**Outputs**:
- Mean Nutrient Concentration
- Pollution Risk as “low”, “med” or “high”

```{r}
source(here("week6/classify_lake.R"))

# generate some data to try the function
temperature = runif(min=4, max=35, n=30)
nutrient = runif(min=5, max=40, n=30)

# try it
res=classify_lake(temperature=temperature, nutrient=nutrient)

# lets pick some numbers we know should give us a low value
nutrient = runif(min=0, max=5, n=30)
classify_lake(temperature=temperature, nutrient=nutrient)
```
### Review

We are using functions as a way to design modular programs (that might involve many functions that are linked together)

We always start with clearly defininng our goal for the whole program and each piece (and diagramming this if needed for programs made of multiple functions)

Always write your function in a text editor and then copy into R

By convention we name files with functions in them by the name of the function.R e.g. power_gen_orig.R

you can have R read a text file by source(“power_gen.R”) - make sure you are in the right working directory

keep organized by keeping all functions in a subdirectory called R

Eventually we will want our function to be part of a package (a library of many functions) - to create a package you must use this convention (name.R) place all function in a directory called “R”


## Error Checking - What will your function do if user gives you garbage data?

imagine that you have a function that computes microbial growth rates with temperature...

**Two Options:**

1. error-checking (nice)

if temperature < -100 or > 100 
- set results to NA
- or exit function with a warning

2. assume user reads the contract :) return unrealistic values
- if input -999.0, will still try to output growth rate

Error-checking is helpful if you are going to build a model made up of many functions- why?

### How to add error checking
```{r}
source(here("week6/power_gen_orig.R"))

power_gen = function(height, flow, rho=1000, g=9.8, Keff=0.8) {
  # make sure inputs are positive and if not set as NA so result will be NA
   if (height < 0) height=NA
  
# an alterative is to exit the function prematurely 
  if (flow < 0) return("Flow cannot be less than 0")
  if (height < 0) return("Height cannot be less than 0")
      
  # calculate power
    result = rho * height * flow * g * Keff
  return(result)
}

power_gen(2,3)
```

