---
title: "Week 4 Notes"
subtitle: "ESM 262 - Computing for Environmental Science and Management"
author: Alecia
format: html
editor: visual
---

# Tuesday 5/5

```{r}
library(tidyverse)
library(here)

```

## Data Types

Some useful data types in R:

- string
- numeric
- tibble
- factors
- lists
- 

## Ways to Generate Data

When testing your code, it can be helpful to generate test data.

Generating numeric data: `runif(min, max)` = sampling from a uniform distribution `rnorm(mean, sd)` = sampling from a uniform and normal distribution `rpois(lambda)` = sampling from a poisson distribution

For categorical data: `sample(vector)` `sequence`

### Examples

- fire starts = uniform distribution

```{r}
firestarts = runif(20, min=0, max = 100)
ggplot(data.frame(firestarts = firestarts),
       aes(firestarts)) + geom_histogram()

```

- otter body weights = normal distribution

```{r}
otterweight = rnorm(100, mean = 10, sd = 2) # first number is n sample size
ggplot(data.frame(otterweight),
       aes(otterweight)) + geom_histogram()
```

Pick a random sample from a vector of values:

```{r}
pick = sample(seq(2000, 2025), size = 20, replace = TRUE)
pick

# no repeat years
pick2 = sample(seq(2000, 2025), size = 20, replace = FALSE)
pick2

```

Bear Encounters:

```{r}
bear = rpois(1000, 1)

ggplot(data.frame(bear = bear),
       aes(bear)) + geom_histogram()
```

```{r}
data()

data(airquality) #grab dataset from R

help(airquality) # learn about data set

str(airquality) #structure of data frame

attributes(airquality) #attributes of data frame

ggplot(airquality, aes(x=Temp, y = Ozone)) + geom_point()

```

Let's say we are testing a new product and we weant a sample of 5 ozone measures (one column of the data)

```{r}
ozone_sample = sample(airquality$Ozone, size = 5, replace = FALSE)
ozone_sample

#results = [1]  7 23 NA 30 91

```

If your data has NA's, then when you take a random sample they may be selected.

If you don't want NA's...

```{r}
airquality_clean = na.omit(airquality) #makes a new data set without any NA's

```

How to sample several ROWS (with all info) from the data set...

```{r}
ozone_sample_full = sample(1:nrow(airquality_clean), size = 5, replace = FALSE)
ozone_sample_full

ggplot(airquality_clean, aes(x = Temp, y = Ozone)) + 
  geom_point() +
  geom_point(aes(x = ozone_sample_full$Temp, y = ozone_sample_full$Ozone), color = "red")
  
```

Wind Speed Practice

```{r}
wind_median <- median(airquality_clean$Wind)

wind_sample = sample(airquality_clean$Wind, size = 10, replace = FALSE)
wind_sample
# [1]  9.2 12.6  8.0  9.7  8.0 12.0  4.6  8.6  8.6  9.2

wind_median_sample = median(wind_sample)

```

## Factors

```{r}
# good for classes, categories. Like if you want to treat "Year" as a factor

# R calls them levels

possible.fish = c("salmon", "steelhead", "shark", "tuna", "cod")

#let's sample from the vector
catch1 = base::sample(possible.fish, size = 20, replace = TRUE)

# because possible,fish was a factor, then catch1 will also be a factor
catch1

summary(catch1) # not that useful...

catch1 = as.factor(catch1)

summary(catch1) #provides each category and how many are in that category
```

Other useful things with factors...

```{r}
plot(catch1, col = "blue") #creates a bar chart with frequencies of categories

mean(summary(catch1)) #7

max(summary(catch1)) #7

which.max(summary(catch1)) # cod 1

names(which.max(summary(catch1))) #just to get the name "cod"
```

Practice = Catch something...

```{r}
# fake population of butterflies

possible.butt = c("orange", "hotpink", "chartreuse", "skyblue", "purple")

catch2 = base::sample(possible.butt, size = 30, replace = TRUE)

catch2 = as.factor(catch2)

summary(catch2)

plot(catch2, col = c("chartreuse","hotpink", "orange", "purple", "skyblue"))
```

## Strings

`sprintf` used something called *format strings*...

**Example:**

```{r}
sprintf("We found %d %s %3.1f miles from shore", 4, "Otters", 18.34985)

# %s replaces with a string
# %d replaces with an integer
# %f replaces with a real value
#4.1f replaces iwth a real value with 4 digits, two after decimal
# %n replaces with a numeric
```

Practice

```{r}
# USE RESULTS FOR CREAING TEXT

```

Make a string with sprintf and add to plot title

```{r}
# # create vector of possible flowers 
possible.flowers = c("tulip","rose","lily","iris","marigold","daisy")

# we can use sample to simulate a random recording of purchases at the market, lets create data that had 300 flower purchases

purchases = sample(possible.flowers, size=20, replace=T)
purchases= as.factor(purchases) # this is my data frame

# make it a data frame for plotting later
# maybe we want to add more information about the purchases later - like price, time of day, etc. - so make column name flower
purchases  = data.frame(flower=purchases) 

# to quickly get information of flower purchases 
summary(purchases$flower)
```

```{r}
# sprintf creates a string %s mean use what ever is after the , to fill in a string
plottitle=sprintf("The most frequently bought flower was %s\n There were %d bought", names(which.max(summary(purchases$flower))), max(summary(purchases$flower)))

# now plot
ggplot(purchases, aes(flower, fill=flower))+geom_histogram(stat="count")+labs(title=plottitle)+
  theme(legend.position="none")
```

# Thursday 5/7

## Looping!

Repeating something is a fundamental concept in programming. We usually call this looping. All programming languages have ways to do this - sometimes hidden. Lets start with the basics of how you do repetition - the *for* loop - used in many languages.

### *for* statement

- defines a loop using a counter that is incremented each time you go through the loop
- the counter - could be any variable (we often use i) but you could use whatever variable you want (usually use some single letter)
- it repeats the commands between the { and } after the *for* keyword and loop definition
- these commands are repeated until the counter condition is TRUE

```{r}
# simple example - use a loop to sum the numbers from 1 to 5
# a = 0+1+2+3+4+5
a=0
for (i in 1:5) {  # the for statement in
 a = a+i
}
a

```

for (i in 1:5) - mean you start with i=1 the commands in the loop are executed - a = a+i then you go back to the top of the loop and increment i to 2

i = 1 a = 0 + 1 = 1 i = 2 a = 1 + 2 = 3 i = 3 a = 3 + 3 = 6

#### Keeping track of what happens each step/interation

- start with a data structure to hold the results -

- initialize to 0

  - its good practice to initialize data before a loop
  - initialize means to create a data structure and set values, usually to zero but sometimes NA

  ```{r}
  # if you want to keep track of a for each iteration
  # start with a data structure to hold the results - 
  # there will be 5 iterations, so a vector of length 5
  # initialize to 0 

  av = rep(x=0, times=5)

  # now we can fill in the vector with the sum at each iteration
  # we can use the counter (i) to access the correct element of the vector
  # note that we have to treat the first iteration differently because we don't have a previous value to add to

  for (i in 1:5) {  # counter is the i
    if (i==1) {     #condition statement if/else
      av[i] = 1
    } else {
    av[i] = av[i-1]+i
    }
  }
  av
  ```

  ```{r}
  # note that R has alternatives using functions that do this looping for you
  a = sum(seq(from=1, to=5))
  a
  ```

### When to use loops

When you want to repeat something a specific number of times R has many built in functions to do this \* BUT sometimes won’t do exactly what you want \* other languages also use versions of the for loop

Generating test data is a helpful way to check your code. Take advantage of built-in R functions to sample from distributions as we did last class

```{r}
testnumbers = runif(min=0, max=20, n=100)

total = 0
nvalues = 100
for (i in 1:nvalues) {
  total = testnumbers[i]+total
}
total/nvalues

```

Check if your loop is correct using `mean`

```{r}
mean(testnumbers)

```

#### Another Example

Calculating net income from revenue minus expenses

```{r}
# lets say we wanted to look at income minus expenses for net income
# create some data for an example

income = runif(min=1000, max=5000, n=10)
income

expenses = rnorm(mean=1500, sd=500, n=10)
expenses

# it is possible that this approach could generate negative expenses, so fix that:
expenses = ifelse(expenses < 0, 0, expenses)

# lets make this a data frame for organization
account = data.frame(income=income, expenses=expenses)
# lets add a year
account$year = seq(from=2000, length.out=nrow(account))

# now compute net income - we could do this with a loop but R is designed to do this without loops
# normally we can just do
account = account %>% mutate(net = income-expenses)
head(account)
```

```{r}
# We could have computed net income with a for loop!
# sometimes we use NA to show we haven't computed it yet, so "initialize" as NA

# lets add a column for our net computed by a loop so our data has somewhere to go
account$netloop=NA
account

# note that I can use any variable I want as a counter -
# notice how I'm now using n to access each row of the data frame
for (n in 1:10) {
  account$netloop[n] = account$income[n]-account$expenses[n]
}

# as expected, mutate net and netloop are the same :)
head(account)
```

### Why loop: cross-row dependencies

While sum and mean are functions created by the R community, what if you were doing something more complex to each row? What if we needed data from a previous row for each calculation? Example: carry-forward net income = current year’s net + last year’s carry-forward

```{r}
# note: we treat the first year differently because we don't have carry forward.
# initialize our column to zero, start counter (n) at 2 instead of 1
# make the first value equal to the net of first year

# create new column
account$carryforwardnet = 0
account$carryforwardnet[1]=account$net[1]
# now loop - looking backward one year to get previous years carry forward
for (n in 2:10) {
  account$carryforwardnet[n] = account$net[n]+account$carryforwardnet[n-1] 
  # n-1 is the previous row
  # have to start counter at 2 because there is no row before the first row
}

# plot - use color to show current year's contribution as positive or negative
ggplot(account, aes(year, carryforwardnet, fill=net, group=year))+geom_col()+
  scale_fill_gradient(low="red", high="black")
```

::: callout-note
When you are adding/subtracting across rows, make sure you think about what happens at the beginning and the end.
:::

### Why loop: sequential comparisons

Another case where loops are necessary: comparing across elements one at a time

Lets say we are measuring **speeds** and we want to determine the number of times speeds are above a threshold

There are many ways to do this - but for will always work

```{r}
# what is our speed limit 
speedthresh = 120

# create some sample speeds
speeds = runif(min=0, max=150, n=300)

nthresh = 0 #create a column for the values from loop to go 
for ( i  in 1:length(speeds)) {   #using length(data) allows you to make sure all rows are counted
   if(speeds[i] > speedthresh)
     nthresh = nthresh + 1
}

nthresh
```

### Add some complexity

Let's imagine that we have data on running speeds of a creature we are trying to develop a conservation plan for.

We think that \* when the creature is running at 35 km/hr and above they are hunting. \* when they are moving below 5 km/hr they are socializing \* when they are moving as speeds between this they are traveling

We want to estimate how often they are in each of these categories based.

Build a for loop to do this.

As before, generate some test data using functions in R that allow you to generate data by sampling from distributions with known parameters

```{r}
speedy = runif(min= 0, max = 50, n = 300)

hunting = 35
social = 5
#initialize the column/place to store the data
num_hunting = 0
num_social = 0

for (i in 1:length(speedy)) {
  num_hunting = num_hunting + ifelse(speedy[i] >= hunting, 1, 0) #condition, value if true, value if false
    num_social = num_social + ifelse(speedy[i] <  social, 1, 0)
}

# other speeds are traveling, is remainder
num_traveling = length(speedy) - num_hunting - num_social

# as a percent of recorded speeds
num_hunting/length(speedy)*100
num_social/length(speedy)*100
num_traveling/length(speedy)*100
```

## Assignment 3

For generated data, shoot for a mean around 20
